# Script to develop  multi-variable prediction model for time-to-event outcomes using
# an internal-external validation framework
# performance metrics include C-index, calibration slope, calibration-in-the-large, Brier score and 
# decision curve analysis

# No data sharing is allowed, so the script generates a simulated dataset 

rm(list=ls())

##
### Load all relevant packages
```{r}
library(tidyverse)
library(survival)
library(dplyr)
library(mice)
library(caret)
library(ggsurvfit)
library(ggplot2)
library(survminer)
library(glmnet)
library(dcurves)
library(pec)
library(rms)

```

##
### Define categorical and  numerical variables
```{r}

# Set seed for reproducibility
set.seed(123)

# Number of observations
n <- 1000

# Define categorical variables
cat_vars <- c("Gender_ID", "ethnicitycleaned", "borough_of_residence", "index_diagnosis",
              "AD_across_rows_6mo", "ANX_across_rows_6mo", "MS_count", "AP_count")

# Define continuous variables
cont_vars <- c("event_merged", "time_to_event", "new_age", "total_length_mh", "total_length_nmh",
               "Positive_symptom_score", "Negative_symptom_score", "Depressive_symptom_score", 
               "Manic_symptom_score", "Disorganised_symptom_score", "Catatonic_symptom_score", 
               "Other_symptoms", "Substance_score")

# Generate simulated data
result <- data.frame(
  brcid_a = sprintf("BRCID-%04d", 1:n),  # Create unique alphanumeric ID codes
  Gender_ID = sample(c("Male", "Female", "Other"), n, replace = TRUE),
  ethnicitycleaned = sample(c("White", "Black", "Asian", "Other"), n, replace = TRUE),
  borough_of_residence = sample(c("Borough_A", "Borough_B", "Borough_C"), n, replace = TRUE),
  index_diagnosis = sample(c("Diagnosis_A", "Diagnosis_B", "Diagnosis_C"), n, replace = TRUE),
  AD_across_rows_6mo = sample(c("Yes", "No"), n, replace = TRUE),
  ANX_across_rows_6mo = sample(c("Yes", "No"), n, replace = TRUE),
  MS_count = sample(0:5, n, replace = TRUE),
  AP_count = sample(0:5, n, replace = TRUE),
  
  event_merged = sample(c(0, 1, 2), n, replace = TRUE),  # Updated to take values 0, 1, or 2
  time_to_event = runif(n, min = 1, max = 5000),
  new_age = rnorm(n, mean = 40, sd = 10),
  total_length_mh = rpois(n, lambda = 50),
  total_length_nmh = rpois(n, lambda = 30),
  Positive_symptom_score = rnorm(n, mean = 10, sd = 3),
  Negative_symptom_score = rnorm(n, mean = 8, sd = 2.5),
  Depressive_symptom_score = rnorm(n, mean = 12, sd = 4),
  Manic_symptom_score = rnorm(n, mean = 9, sd = 3),
  Disorganised_symptom_score = rnorm(n, mean = 7, sd = 2),
  Catatonic_symptom_score = rnorm(n, mean = 5, sd = 1.5),
  Other_symptoms = rnorm(n, mean = 6, sd = 2),
  Substance_score = rpois(n, lambda = 3)
)

# Convert categorical variables to factors
result[cat_vars] <- lapply(result[cat_vars], factor)

# Convert continuous variables to numeric (ensures correct type)
result[cont_vars] <- lapply(result[cont_vars], as.numeric)

# Introduce missing values at random
introduce_missingness <- function(data, vars_to_exclude, missing_prob) {
  # Get all variable names except the ones to exclude
  vars <- setdiff(names(data), vars_to_exclude)
  
  for (var in vars) {
    missing_indices <- sample(1:nrow(data), size = round(missing_prob * nrow(data)), replace = FALSE)
    data[missing_indices, var] <- NA
  }
  return(data)
}

# Variables to exclude from missingness
vars_to_exclude <- c("event_merged", "time_to_event")

# Define missingness probability
missing_prob <- 0.1

# Apply the function to the dataset
result <- introduce_missingness(result, vars_to_exclude, missing_prob)

# Inspect the dataset
str(result)

# View a sample of the data with missing values
head(result)

```

##
### Check missingness
```{r}

# Double check how much missing data there is for the predictors
na_count <- sapply(result, function(x) sum(is.na(x)))
na_count

```

##
### Censor at 6 years 
```{r}


censoring_threshold <- 6 * 365  # 6 years in days
result$cens <- ifelse(result$time_to_event > censoring_threshold, 1, 0)  # Censoring indicator
result$cens_event <- ifelse(result$time_to_event > censoring_threshold, 0, as.character(result$event_merged))  # Censored event
result$cens_time <- pmin(result$time_to_event, censoring_threshold)  # Censored time
max(result$cens_time, na.rm = TRUE)

```

##
### Merge sperate events 
```{r}
                   
result$event_merged <- result$cens_event  %>% as.numeric() 
result$event_merged <- ifelse(result$event_merged == 1 | result$event_merged == 2, 1, result$event_merged)
result$time_to_event <- result$cens_time  %>% as.numeric() 
result <- result %>% select(-cens_event, -cens_time, -cens)
```


##
### Basic survival curve
```{r}


surv_obj <- Surv(time = result$time_to_event, event = result$event_merged) # Event should be numeric otherwise it assumes a multi-state model (e.g. competing risks)

surv_fit <- survfit(surv_obj ~ 1)

time_points <- c(0,365,2*265,3*365,4*365,5*365,6*365)
surv_summary <- summary(surv_fit, times = time_points )

summary_table <- data.frame( 
    Time = time_points, 
    Cumulative_Risk = signif(1 - surv_summary$surv, 3), 
    Lower_CI = signif(1 - surv_summary$lower, 3), 
    Upper_CI = signif(1 - surv_summary$upper, 3), 
    Number_events =  signif(surv_summary$n.event, 3), 
    Sample_size = signif(surv_summary$n.risk, 3)
    )

print(summary_table)

#write.csv(summary_table, "survival_prob_joint.csv")

```
##
### Plot cumulative risk
```{r}

# Convert the survfit object to a data frame
surv_data <- data.frame(
  time = surv_fit$time,
  prob_event = 1 - surv_fit$surv,  # Complement of survival probability
  lower = 1 - surv_fit$lower,      # Lower confidence interval
  upper = 1 - surv_fit$upper       # Upper confidence interval
)

# Create the step plot with confidence intervals using ggplot
ggplot(surv_data, aes(x = time, y = prob_event)) +
theme_minimal() +
  # Step plot for the event probability
  geom_step(color = "black", size = 1) + 
  
  # Add the confidence interval as a ribbon (shaded area)
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray", alpha = 0.5, color = NA) +
  
  # Optional: Add step lines for the lower and upper confidence intervals
  # geom_step(aes(y = lower), color = "blue", linetype = "dashed", size = 0.5) +  # Lower CI
  # geom_step(aes(y = upper), color = "blue", linetype = "dashed", size = 0.5) +  # Upper CI
  
  # Labels and title
  labs(x = "Time (days)", y = "Probability of Event Occurring", 
       title = "") +

 # Custom X-axis scale with breaks at intervals of 365 and labels matching those breaks
  scale_x_continuous(breaks = seq(0, 2190, by = 365), labels = seq(0, 2190, by = 365)) +
  
  
  # Y-axis scale between 0 and 1 for probabilities
  scale_y_continuous(limits = c(0, 0.4), expand = c(0, 0)) +
  
  # Apply minimal theme for a clean look
  theme_minimal() +
  
  # Customize text and title sizes
  theme(
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
    axis.text = element_text(size = 17),
    axis.title = element_text(size = 20)
      
  )

```

##
### Build the final model first on all the available data
```{r}

# Imputation --------------

# Remove event and time-to-event from imputation

outcome <- result %>% select(c("time_to_event","event_merged"))

result_no_outcome <- result %>% select(-c("brcid_a","time_to_event","event_merged"))
all_vars <- colnames(result_no_outcome)
set.seed(123)
mice_imputes <- suppressMessages(mice(result_no_outcome, method = "rf", printFlag = FALSE)) # Mice with Random forest 

# Check the summary 
summary(mice_imputes)
```


##
### Complete the imputed data
```{r}

imputed_data <- complete(mice_imputes)
imputed_data <- cbind(imputed_data, outcome)

head(imputed_data)

sum(is.na(imputed_data$new_age))
sum(is.na(imputed_data$Gender_ID))
sum(is.na(imputed_data$ethnicitycleaned))

```

##
### Now fit full model
```{r}

x_full <- imputed_data %>% select(all_of(all_vars))

# Dummify the data
dmy <- dummyVars(" ~ .", data = x_full, fullRank = TRUE)
trsf_full <- data.frame(predict(dmy, newdata = x_full))
x_full <- as.matrix(trsf_full)

y_full <- Surv(imputed_data$time_to_event, imputed_data$event_merged)

```

```{r}
# Fit the Lasso-Cox model on the full data

final_model <- cv.glmnet(x_full, y_full, alpha=1, family="cox", type.measure = "C", nfolds = 5)

# Extract the lambda values
lambda_values <- final_model$lambda                   

# Extract the C-index values (cv-means represent the performance)
c_index_values <- final_model$cvm

# Extract the standard errors for the C-index
c_index_se <- final_model$cvsd

# Plot the C-index values vs. lambda
plot(lambda_values, 
     c_index_values, 
     type = "b",               # Line and point plot
     log = "x",                # Plot lambda on a log scale
     xlab = "Lambda (log scale)", 
     ylab = "C-Index", 
     main = "C-Index vs. Lambda",
     col = "blue",             # Line color
     lty = 1,                  # Line type
     pch = 19)                 # Point type

# Add error bars for the C-index (±1 standard error)
arrows(lambda_values, 
       c_index_values - c_index_se, 
       lambda_values, 
       c_index_values + c_index_se, 
       length = 0.05, 
       angle = 90, 
       code = 3, 
       col = "blue")

# Add grid for better visualization
grid()


# Extract the lambda that gives the minimum cross-validation error (highest C-index)
best_lambda_min <- final_model$lambda.min
cat("Best lambda (lambda.min):", best_lambda_min, "\n")
```


```{r}
# Extract the coefficients at the optimal lambda
final_coefficients <- as.matrix(coef(final_model, s = "lambda.min")) %>% as.data.frame()

# Add column for hazard ration (HR)
final_coefficients$'1' <- as.numeric(final_coefficients$'1' ) %>% round(3)
final_coefficients$HR <- exp(final_coefficients$'1' ) %>% round(3)
colnames(final_coefficients) <- c("Coefficient", "HR")
print(final_coefficients)

#write.csv(final_coefficients, "final_coefficients.csv")

```


# Internal-external validation framework
```{r}

list_boroughs <- levels(result$borough_of_residence)
print(list_boroughs)


c_stat_app <- data.frame(C_Cox = rep(NA, length(list_boroughs)),
                         SE_Cox = rep(NA, length(list_boroughs)),
                         Val_borough = rep(NA, length(list_boroughs)),
                         n_train = rep(NA, length(list_boroughs)),
                         events_train = rep(NA, length(list_boroughs)),
                         n_test = rep(NA, length(list_boroughs)),
                         events_test = rep(NA, length(list_boroughs)))

c_stat_ext <- data.frame(C_Cox = rep(NA, length(list_boroughs)),
                         SE_Cox = rep(NA, length(list_boroughs)),
                         Val_borough = rep(NA, length(list_boroughs)),
                         n_train = rep(NA, length(list_boroughs)),
                         events_train = rep(NA, length(list_boroughs)),
                         n_test = rep(NA, length(list_boroughs)),
                         events_test = rep(NA, length(list_boroughs)),
                         cali_1 = rep(NA, length(list_boroughs)),
                         cali_2 = rep(NA, length(list_boroughs)),
                         brier = rep(NA, length(list_boroughs)))
```

```{r}
for (i in 1:length(list_boroughs)) {
 
borough <- list_boroughs[[i]] # need to repeat this n times for each borough 
print(borough)

# Split datasets by borough (leave one out for validation, remaining ones for dvlpt)

train <- result %>% filter(borough_of_residence != borough)  
test <- result %>% filter(borough_of_residence == borough)

# Imputation seperately ------------------------------------------------------------------------------
set.seed(123 + i)

    # Remove event and time-to-event from imputation

outcome_train <- train %>% select(c("time_to_event","event_merged"))
train <- train %>% select(-c("time_to_event","event_merged"))
    
outcome_test <- test %>% select(c("time_to_event","event_merged"))
test <- test %>% select(-c("time_to_event","event_merged"))

    # imputation
mice_imputes_train <- suppressMessages(
  mice(train, method = "rf", printFlag = FALSE)) # Mice with Random forest 
train <- complete(mice_imputes_train)

mice_imputes_test <- suppressMessages(
  mice(test, method = "rf", printFlag = FALSE)) # Mice with Random forest 
test <- complete(mice_imputes_test)

        # merge outcomes after imputation
train <- cbind(train, outcome_train)
train$time_to_event <- as.numeric(train$time_to_event)
train$event_merged <- as.numeric(train$event_merged)

test <- cbind(test, outcome_test)
test$time_to_event <- as.numeric(test$time_to_event)
test$event_merged <- as.numeric(test$event_merged)

    # Add interaction term to the data
train$num_gender <- as.numeric(train$Gender_ID)
train$age_gender_interaction <- train$new_age * train$num_gender

test$num_gender <- as.numeric(test$Gender_ID)
test$age_gender_interaction <- test$new_age * test$num_gender

# Inner fold CV --------------------------------------------------------------------------

  # Prepare the training data for modeling
  x <- train %>% select(all_of(all_vars))
  # dummify the data
  dmy <- dummyVars(" ~ .", data = x, fullRank=T)
  trsf <- data.frame(predict(dmy, newdata = x))
  x <- as.matrix(trsf)
  
  # Check for NA/NaN/Inf in the predictor matrix
  if (any(!is.finite(x))) {
    stop("NA/NaN/Inf values found in the predictor matrix for training data")
  }
  
  y <- Surv(train$time_to_event, train$event_merged)
  
  # Fit the Lasso-Cox model using cross-validation
  train.m <- cv.glmnet(x, y, family="cox", type.measure = "C", nfolds=5)
  
  # Predict the linear predictors (PI) from the Lasso-Cox model
  coef_lasso <- as.data.frame(predict(train.m, s = "lambda.min", newx = x, type = "link", exact = TRUE))
  train <- train %>% mutate(PI = as.matrix(coef_lasso)) 
  
  # Fit a Cox model on the penalized coefficients using coxph to extract the estimates
  cox_model_train <- coxph(Surv(time_to_event, event_merged) ~ PI , data = train)
    print(summary(cox_model_train))   # coefficeitn should be 1 (perfect calibration on train data)

 # Store apparent performance metrics (discrimination, could also output calibration)
  
  c_stat_app[i,1] <- (summary(cox_model_train)$concordance)[1] #%>% round(3)
  c_stat_app[i,2] <- (summary(cox_model_train)$concordance)[2] #%>% round(3)
  c_stat_app[i,3] <- borough
  c_stat_app[i,4] <- nrow(train)
  c_stat_app[i,5] <- sum(train$event_merged)
  c_stat_app[i,6] <- nrow(test)
  c_stat_app[i,7] <- sum(test$event_merged)

# Outer fold test-----------------------------------------------------------------------------------------
  
  # Prepare the test data for validation
  x_test <- test %>% select(all_of(all_vars))
  # dummify the data
  dmy <- dummyVars(" ~ .", data = x_test, fullRank = T)
  trsf <- data.frame(predict(dmy, newdata = x_test))
  x <- as.matrix(trsf)
  x_test <- as.matrix(x)
    
  y_test <- test %>% select(c(event_merged, time_to_event))
  
  # Check for NA/NaN/Inf in the predictor matrix
  if (any(!is.finite(x_test))) {
    stop("NA/NaN/Inf values found in the predictor matrix for test data")
  }

 # Predict the linear predictors (PI) from the Lasso-Cox model on the test set
  coef_lasso_test <- as.data.frame(predict(train.m, s = "lambda.min", newx = x_test, type = "link", exact = TRUE))
  test <- test %>% mutate(PI = as.matrix(coef_lasso_test)) 
  cox_model_test <- coxph(Surv(time_to_event, event_merged) ~ PI , data = test, x= TRUE)

 # Calibration

 # Option 1
 summ <- summary(cox_model_test)
 cali_1 <- summ$coefficients[1]

 # Option 2
     x_test <- as.data.frame(x_test)
     times <- c(2190) # 6 year prediction
     y_surv <- 1 - with(test, Surv(time_to_event, event_merged))[,2]
     surv_prob <- pec::predictSurvProb(cox_model_test, test, times)
     calibration <- val.prob(p=surv_prob, y = y_surv, m=200, pl=F)

     brier <- unname(calibration[11])
     cali_2 <- unname(calibration[13])

    # DCA
    dca_pred_obs <- cbind(y_test,surv_prob) %>% as.data.frame() 
    names(dca_pred_obs)[3] <- "pred"
    dca_pred_obs$pred <- 1- dca_pred_obs$pred # probability of event = 1- surv prob
    #dca_pred_obs$obs <- ifelse(dca_pred_obs$event_merged == 1 & dca_pred_obs$time_to_event < 2190, 1, 0) 
    write.csv(dca_pred_obs, paste0("dca_pred_obs",i ,".csv"), row.names = FALSE) 
    
    dca_assessment  <-  dca(Surv(time_to_event,event_merged) ~ pred,
                        data = dca_pred_obs,
                         time = 2190,
                        thresholds = seq(0, 1, 0.01)) %>% as_tibble() 
    # Summarize net benefit
    dca_assessment = dca_assessment %>%
    group_by(variable, label, threshold)
    
    write.csv(dca_assessment, paste0("dca_assessment",i ,".csv"), row.names = FALSE) 

  # Store external performance metrics
  
  c_stat_ext[i,1] <- (summary(cox_model_test)$concordance)[1] 
  c_stat_ext[i,2] <- (summary(cox_model_test)$concordance)[2]
  c_stat_ext[i,3] <- borough
  c_stat_ext[i,4] <- nrow(train)
  c_stat_ext[i,5] <- sum(train$event_merged)
  c_stat_ext[i,6] <- nrow(test)
  c_stat_ext[i,7] <- sum(test$event_merged)
  c_stat_ext[i,8] <- cali_1  %>% round(3)
  c_stat_ext[i,9] <- cali_2  %>% round(3)
  c_stat_ext[i,10] <- brier 

  
}

```
### Print and save results
 
 
```{r}

print(c_stat_app)
print(c_stat_ext)
mean(c_stat_app$C_Cox)
mean(c_stat_ext$C_Cox)

write.csv(c_stat_app, "c_stat_app.csv")
write.csv(c_stat_ext, "c_stat_ext.csv") 
 
```



### Plots DCA curve
 
 
```{r}

# Then plot DCA function

csv_list1 <- read.csv("Downloads/dca_assessment1.csv")
csv_list2 <- read.csv("Downloads/dca_assessment2.csv")
csv_list3 <- read.csv("Downloads/dca_assessment3.csv")

dca_all <- rbind(csv_list1, csv_list2) 
dca_all <- rbind(dca_all, csv_list3) 
head(dca_all)

# Then plot DCA function

ggplot(data=dca_all, aes(x = threshold, y = net_benefit, color = label)) +
stat_smooth(method = "loess", se = TRUE, formula = "y ~ x") +
coord_cartesian(ylim = c(-0.005, 1)) +
scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits=c(0,0.5)) +
labs(x = "Threshold Probability", y = "Net Benefit", color = "") +
 theme_classic() +
theme(legend.title = element_text(size=20),
      legend.text = element_text(size = 15),
      axis.text = element_text(size = 15),       # Enlarges tick labels
      axis.title = element_text(size = 25),
text = element_text( size =30)) 

 # ggsave("overall_DCA_summary.png", height =5, width =5 )  # scale=0.5


```
  


